\mytitle

### 2021-12-20 paper.tex
fix typos

As is well known, the performante of this naive implementation can be
improved by applying Horner's scheme, which reduces the number of
expensive floating\-/point multiplications to 7 and requires 6
additions:
%
\begin{align}
\begin{split}
  \ln(1+x) \approx{}& x \bigg\{1 + x \bigg[-\frac{1}{2} + x \bigg(\frac{1}{3} + x \bigg\{-\frac{1}{4} \\
  &+ x \bigg[\frac{1}{5} + x\bigg(-\frac{1}{6} + \frac{x}{7}\bigg)\bigg]\bigg\}\bigg)\bigg]\bigg\}.
\end{split}\label{eq:horner}%
\end{align}
%
A C implementation using Horner's scheme could read
%
\lstinputlisting[language=C]{src/log1p_horner.c}
%
Note, that the formulation using Horner's scheme requires the
evaluation of the floating\-/point operations in strictly sequential
order.  However, the significant reduction of the number of
floating\-/point multiplications, compared to the naive
implementation, usually results in a shorter run\-/time.

@@

We define the real dilogarithm $\Li(x)$ for all $x\in\mathbb{R}$ as
the real part of its complex continuation.  The real dilogarithm can
be mapped onto the interval $[0,1/2]$ by using the following
identities \cite{lewin}:
%
\begin{align}
  \Li(x) ={}& \ln(1-x)\left[\frac{1}{2}\ln(1-x) - \ln(-x)\right] - \frac{\pi^2}{6}
              \notag \\ &+\Li\left(\frac{1}{1-x}\right), \quad x\leq -1, \\
  \Li(x) ={}& -\Li\left(\frac{x}{x-1}\right) - \frac{1}{2}\ln^2(1-x),
              \notag \\ & -1<x<0, \\
  \Li(x) ={}& -\Li(1-x) + \frac{\pi^2}{6} - \ln(x) \ln(1-x),
              \notag \\ &\frac{1}{2}<x<1, \\
  \Li(x) ={}& \frac{\pi^2}{6} - \ln(x)\left[\ln\left(1-\frac{1}{x}\right) + \frac{1}{2}\ln(x)\right]
              \notag \\ &+ \Li\left(1-\frac{1}{x}\right), \quad 1<x\leq 2, \\
  \Li(x) ={}& -\Li\left(\frac{1}{x}\right) + \frac{\pi^2}{3} - \frac{1}{2} \ln^2(x), \quad x>2,
\end{align}
%
and by using the special values $\Li(-1)=-\pi^2/12$, $\Li(0)=0$,
$\Li(1/2)=\pi^2/12-\ln^2(2)/2$, $\Li(1)=\pi^2/6$ and $\Li(2)=\pi^2/4$.
Within the interval $[0,1/2]$ the real dilogarithm has the series
expansion \eqref{eq:series}, which can be written as
%
\begin{align}
  \Li(x) = x \sum_{k=0}^\infty \frac{x^{k}}{(k+1)^2}.
  \label{eq:series2}
\end{align}
%
We approximate the sum on the r.h.s.\ of Eq.~\eqref{eq:series2} in the
form of a rational minimax approximant on the interval $[0,1/2]$ using
the \texttt{MiniMaxApproximation} from Wolfram/Mathematica
\cite{mathematica},
%
\begin{align}
  \sum_{k=0}^\infty \frac{x^{k}}{(k+1)^2} \approx
  \frac{\sum_{k=0}^5 p_kx^k}{\sum_{k=0}^6 q_kx^k}.
  \label{eq:minimax}
\end{align}
%
The coefficients $p_k$ and $q_k$ are listed in \tabref{tab:coeffs}.
The approximant on the r.h.s.\ of Eq.~\eqref{eq:minimax} has a maximum
relative error of around $5\cdot 10^{-17}$, suited for an evaluation
of $\Li(x)$ with double precision.
%
\begin{table}[t]
  \centering
  \caption{Coefficients of the numerator and denominator polynomials
    for the minimax approximant \eqref{eq:minimax}.}
  \begin{tabular}{lr}
    \toprule
    $p_0$ & $ 0.9999999999999999502\cdot 10^{+0}$ \\
    $p_1$ & $-2.6883926818565423430\cdot 10^{+0}$ \\
    $p_2$ & $ 2.6477222699473109692\cdot 10^{+0}$ \\
    $p_3$ & $-1.1538559607887416355\cdot 10^{+0}$ \\
    $p_4$ & $ 2.0886077795020607837\cdot 10^{-1}$ \\
    $p_5$ & $-1.0859777134152463084\cdot 10^{-2}$ \\
    $q_0$ & $ 1.0000000000000000000\cdot 10^{+0}$ \\
    $q_1$ & $-2.9383926818565635485\cdot 10^{+0}$ \\
    $q_2$ & $ 3.2712093293018635389\cdot 10^{+0}$ \\
    $q_3$ & $-1.7076702173954289421\cdot 10^{+0}$ \\
    $q_4$ & $ 4.1596017228400603836\cdot 10^{-1}$ \\
    $q_5$ & $-3.9801343754084482956\cdot 10^{-2}$ \\
    $q_6$ & $ 8.2743668974466659035\cdot 10^{-4}$ \\
    \bottomrule
  \end{tabular}
  \label{tab:coeffs}
\end{table}
%
Note that the approximant on the r.h.s.\ of Eq.~\eqref{eq:minimax} is
very similar to the results presented in Ref.~\cite{morris} (DILOG
0011), although with slightly different coefficients.  In oder to
allow for the use of ILP for the evaluation of the numerator and the
denominator we use Estrin's scheme,
%
\begin{align}
  \Li(x) &\approx x\frac{P(x)}{Q(x)}, \qquad 0\leq x \leq \frac{1}{2},
  \label{eq:final}
\end{align}
%
with
%
\begin{align}
  P(x) &= p_0 + x p_1 + y (p_2 + x p_3) + z (p_4 + x p_5), \\
  Q(x) &= q_0 + x q_1 + y (q_2 + x q_3) + z (q_4 + x q_5 + y q_6),
\end{align}
%
where $y=x^2$ and $z=y^2$.  A C implementation of Eq.~\eqref{eq:final}
can be found in the appendix.

@@

In \tabref{tab:runtime} the run\-/time of different C implementations is
compared to the implementation given in the appednix on two different
CPU architectures.  On each architecture the implentations are
compiled with \texttt{g++} with optimization \texttt{-O2}.  The shown
run\-/time denotes the total wall clock time for the calculation of
$\Li(x)$ for $10^6$ uniformly distributed random values $x\in[0,1/2]$.
The choice of this interval ensures that the measurement is only
sensitive to the formulation of the approximant, but not to the
transformation onto the interval $[0,1/2]$.%
%
\begin{table}[tb]
  \centering
  \caption{Run\-/time in seconds for the calculation of $\Li(x)$ for
    different implementations on different CPU architectures with
    different compilers.}
  \begin{tabular}{lcc}
    \toprule
    \multirow{2}{*}{Implementation} & i7-4700MQ & i7-5600U \\
    & \texttt{g++} 9.3.0 & \texttt{g++} 10.2.1 \\
    \midrule
    Eq.~\eqref{eq:final} (this paper) & $0.0076$ & $0.0058$ \\
    Morris (DILOG 0011) \cite{morris} & $0.0115$ & $0.0085$ \\
    Cephes \cite{cephes}              & $0.0136$ & $0.0119$ \\
    Algorithm 327 \cite{koelbigDilog} & $0.0229$ & $0.0172$ \\
    Chebyshev \cite{luke,root}        & $0.0718$ & $0.0515$ \\
    Algorithm 490 \cite{ginsberg}     & $0.1057$ & $0.0576$ \\
    GSL \cite{gsl}                    & $0.2091$ & $0.0898$ \\
    \bottomrule
  \end{tabular}
  \label{tab:runtime}
\end{table}%

@@

The third fastest implementation is provided by Cephes \cite{cephes},
which implements a rational function approximation for Spence's
function $\operatorname{Sp}(x)=\Li(1-x)$.  The implementation provided
by Cephes is by a factor two slower than the implementation prestend
in this paper.  This longer run\-/time has several reasons: For one,
Cephes uses a rational function approximation with more terms in the
numerator and denominator.  Furthermore, Cephes evaluates the
numerator and denominator polynomials using Horner's scheme, which is
purely sequential.\footnote{For $x>1$ Cephes calculates $\ln(x)$
  twice, which leads to a further increased run\-/time when a benchmark
  interval with $x>1$ is chosen.}

@@

A representation in terms of Chebyshev polynomials \cite{luke} is
implemented in the ROOT program library \cite{root}.  In this
implementation the polynomials are summed using Clenshaw's algorithm,
which is also purely sequencial and slightly more complex than
Horner's scheme.  This is a reason why the implementation in
\cite{root} has a larger run\-/time than the rational function
approximants discussed above.

>>>>

As is well known, the performance of this naive implementation can be
improved by applying Horner's scheme, which reduces the number of
expensive floating\-/point multiplications to 7 and requires 6
additions:
%
\begin{align}
\begin{split}
  \ln(1+x) \approx{}& x \bigg\{1 + x \bigg[-\frac{1}{2} + x \bigg(\frac{1}{3} + x \bigg\{-\frac{1}{4} \\
  &+ x \bigg[\frac{1}{5} + x\bigg(-\frac{1}{6} + \frac{x}{7}\bigg)\bigg]\bigg\}\bigg)\bigg]\bigg\}.
\end{split}\label{eq:horner}%
\end{align}
%
A C implementation using Horner's scheme could read
%
\lstinputlisting[language=C]{src/log1p_horner.c}
%
Note, that the formulation using Horner's scheme requires the
evaluation of the floating\-/point operations in strictly sequential
order.  However, the significant reduction of the number of
floating\-/point multiplications, compared to the naive
implementation, usually results in a shorter run\-/time.

@@

We define the real dilogarithm $\Li(x)$ for all $x\in\mathbb{R}$ as
the real part of its complex continuation.  The real dilogarithm can
be mapped onto the interval $[0,1/2]$ by using the following
identities \cite{lewin}:
%
\begin{align}
  \Li(x) ={}& \ln(1-x)\left[\frac{1}{2}\ln(1-x) - \ln(-x)\right] - \frac{\pi^2}{6}
              \notag \\ &+\Li\left(\frac{1}{1-x}\right), \quad x\leq -1, \\
  \Li(x) ={}& -\Li\left(\frac{x}{x-1}\right) - \frac{1}{2}\ln^2(1-x),
              \notag \\ & -1<x<0, \\
  \Li(x) ={}& -\Li(1-x) + \frac{\pi^2}{6} - \ln(x) \ln(1-x),
              \notag \\ &\frac{1}{2}<x<1, \\
  \Li(x) ={}& \frac{\pi^2}{6} - \ln(x)\left[\ln\left(1-\frac{1}{x}\right) + \frac{1}{2}\ln(x)\right]
              \notag \\ &+ \Li\left(1-\frac{1}{x}\right), \quad 1<x\leq 2, \\
  \Li(x) ={}& -\Li\left(\frac{1}{x}\right) + \frac{\pi^2}{3} - \frac{1}{2} \ln^2(x), \quad x>2,
\end{align}
%
and by using the special values $\Li(-1)=-\pi^2/12$, $\Li(0)=0$,
$\Li(1/2)=\pi^2/12-\ln^2(2)/2$, $\Li(1)=\pi^2/6$ and $\Li(2)=\pi^2/4$.
Within the interval $[0,1/2]$ the real dilogarithm has the series
expansion \eqref{eq:series}, which can be written as
%
\begin{align}
  \Li(x) = x \sum_{k=0}^\infty \frac{x^{k}}{(k+1)^2}.
  \label{eq:series2}
\end{align}
%
We approximate the sum on the r.h.s.\ of Eq.~\eqref{eq:series2} in the
form of a rational minimax approximant on the interval $[0,1/2]$ using
the \texttt{MiniMaxApproximation} from Wolfram/Mathematica
\cite{mathematica},
%
\begin{align}
  \sum_{k=0}^\infty \frac{x^{k}}{(k+1)^2} \approx
  \frac{\sum_{k=0}^5 p_kx^k}{\sum_{k=0}^6 q_kx^k}.
  \label{eq:minimax}
\end{align}
%
The coefficients $p_k$ and $q_k$ are listed in \tabref{tab:coeffs}.
The approximant on the r.h.s.\ of Eq.~\eqref{eq:minimax} has a maximum
relative error of around $5\cdot 10^{-17}$, suited for an evaluation
of $\Li(x)$ with double precision.
%
\begin{table}[t]
  \centering
  \caption{Coefficients of the numerator and denominator polynomials
    for the minimax approximant \eqref{eq:minimax}.}
  \begin{tabular}{lr}
    \toprule
    $p_0$ & $ 0.9999999999999999502\cdot 10^{+0}$ \\
    $p_1$ & $-2.6883926818565423430\cdot 10^{+0}$ \\
    $p_2$ & $ 2.6477222699473109692\cdot 10^{+0}$ \\
    $p_3$ & $-1.1538559607887416355\cdot 10^{+0}$ \\
    $p_4$ & $ 2.0886077795020607837\cdot 10^{-1}$ \\
    $p_5$ & $-1.0859777134152463084\cdot 10^{-2}$ \\
    $q_0$ & $ 1.0000000000000000000\cdot 10^{+0}$ \\
    $q_1$ & $-2.9383926818565635485\cdot 10^{+0}$ \\
    $q_2$ & $ 3.2712093293018635389\cdot 10^{+0}$ \\
    $q_3$ & $-1.7076702173954289421\cdot 10^{+0}$ \\
    $q_4$ & $ 4.1596017228400603836\cdot 10^{-1}$ \\
    $q_5$ & $-3.9801343754084482956\cdot 10^{-2}$ \\
    $q_6$ & $ 8.2743668974466659035\cdot 10^{-4}$ \\
    \bottomrule
  \end{tabular}
  \label{tab:coeffs}
\end{table}
%
Note that the approximant on the r.h.s.\ of Eq.~\eqref{eq:minimax} is
very similar to the results presented in Ref.~\cite{morris} (DILOG
0011), although with slightly different coefficients.  In order to
allow for the use of ILP for the evaluation of the numerator and the
denominator we use Estrin's scheme,
%
\begin{align}
  \Li(x) &\approx x\frac{P(x)}{Q(x)}, \qquad 0\leq x \leq \frac{1}{2},
  \label{eq:final}
\end{align}
%
with
%
\begin{align}
  P(x) &= p_0 + x p_1 + y (p_2 + x p_3) + z (p_4 + x p_5), \\
  Q(x) &= q_0 + x q_1 + y (q_2 + x q_3) + z (q_4 + x q_5 + y q_6),
\end{align}
%
where $y=x^2$ and $z=y^2$.  A C implementation of Eq.~\eqref{eq:final}
can be found in the appendix.

@@

In \tabref{tab:runtime} the run\-/time of different C implementations is
compared to the implementation given in the appendix on two different
CPU architectures.  On each architecture the implementations are
compiled with \texttt{g++} with optimization \texttt{-O2}.  The shown
run\-/time denotes the total wall clock time for the calculation of
$\Li(x)$ for $10^6$ uniformly distributed random values $x\in[0,1/2]$.
The choice of this interval ensures that the measurement is only
sensitive to the formulation of the approximant, but not to the
transformation onto the interval $[0,1/2]$.%
%
\begin{table}[tb]
  \centering
  \caption{Run\-/time in seconds for the calculation of $\Li(x)$ for
    different implementations on different CPU architectures with
    different compilers.}
  \begin{tabular}{lcc}
    \toprule
    \multirow{2}{*}{Implementation} & i7-4700MQ & i7-5600U \\
    & \texttt{g++} 9.3.0 & \texttt{g++} 10.2.1 \\
    \midrule
    Eq.~\eqref{eq:final} (this paper) & $0.0076$ & $0.0058$ \\
    Morris (DILOG 0011) \cite{morris} & $0.0115$ & $0.0085$ \\
    Cephes \cite{cephes}              & $0.0136$ & $0.0119$ \\
    Algorithm 327 \cite{koelbigDilog} & $0.0229$ & $0.0172$ \\
    Chebyshev \cite{luke,root}        & $0.0718$ & $0.0515$ \\
    Algorithm 490 \cite{ginsberg}     & $0.1057$ & $0.0576$ \\
    GSL \cite{gsl}                    & $0.2091$ & $0.0898$ \\
    \bottomrule
  \end{tabular}
  \label{tab:runtime}
\end{table}%

@@

The third fastest implementation is provided by Cephes \cite{cephes},
which implements a rational function approximation for Spence's
function $\operatorname{Sp}(x)=\Li(1-x)$.  The implementation provided
by Cephes is by a factor two slower than the implementation presented
in this paper.  This longer run\-/time has several reasons: For one,
Cephes uses a rational function approximation with more terms in the
numerator and denominator.  Furthermore, Cephes evaluates the
numerator and denominator polynomials using Horner's scheme, which is
purely sequential.\footnote{For $x>1$ Cephes calculates $\ln(x)$
  twice, which leads to a further increased run\-/time when a benchmark
  interval with $x>1$ is chosen.}

@@

A representation in terms of Chebyshev polynomials \cite{luke} is
implemented in the ROOT program library \cite{root}.  In this
implementation the polynomials are summed using Clenshaw's algorithm,
which is also purely sequential and slightly more complex than
Horner's scheme.  This is a reason why the implementation in
\cite{root} has a larger run\-/time than the rational function
approximants discussed above.
